How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-09-03 analysis/paper/paper.Rmd
Editing the manuscipt based on Reviewer2's comments

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artifacts are our only window into these behaviors. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around ca. 32 ka, contains fauna specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to ca. 29 ka, there is a small scatter of stone artifacts consisting of blades, Levallois flakes, cores and other retouched flakes. Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Shuidonggou shows that although stemmed points are not present at all, a blade technology came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9). Advanced core and flake tools were likely to have been locally developed at 33 ka (SDG Locality 2 and 8), and at 10.8 ka microblade technology appears, but the origin is uncertain (SDG Locality 12) [@yi2014significance; @li2019history]. The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards the Last Glacial Maximum (LGM) [@yi2014significance].

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are Chert and Andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

Figure \@ref(fig:volume-density) shows that there is no strong directional temporal pattern of changes in artifact density. The upper bound of artifact density increases in more recent periods. For example, the density of some assemblages are two or three times higher than others, especially after 30 ka. The inset figure shows that assemblages with stemmed points have higher artifact densities than the assemblages without. With the newer tools, hunter-gatherers may have stayed in the same location for longer durations. We may be seeing site occupation patterns in this sample that represent the long term trend of the use of both base camps or residences, represented by the artifact higher densities, and more briefly occupied camps with lower densities. The most striking result here is that maximum size of base camps increases in more recent periods.

@@

Previous work on toolkit composition during the Late Pleistocene of Korea attempted to divide it up into multiple chronological sub-periods. Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artifacts; (2) mostly small quartzite and quartz vein artifacts with some large artifacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30 ka [@seong2015diversity]. Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; 'core tool industry with large flake knapping' during early MIS 3 (59−40 ka), and 'blade industry with tanged point chiefly' during late MIS (40−24 ka) [@lee2012characteristics]. Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement. Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, support the continuation of core tools abundantly throughout the Late Pleistocene. In addition, our data show that side scrapers remained as major tools while other tools occupied only a small portion of the assemblages. As stemmed points and blades increased in the toolkits, both choppers and side scrapers went through minor decreases but still remained abundant.

>>>>

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most open sites in Korea do not preserve organic remains due to acidic soil. There are a few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts are our only window to examine hunting and subsistence strategies. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @bae2017late]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, ground stone tools etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean]. Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2012nature; @bae2017origin]. For example, the new dating results from Zhoukoudian Upper Cave from northern China (35.1–33.5 ka) and material evidence such as ornaments and hominin fossil show close relation to one of multiple dispersal events out of Africa into Eurasia    [@li2018re].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around ca. 32 ka, contains faunal specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to ca. 29 ka, there is a small scatter of stone artifacts consisting of blades, Levallois flakes, cores and other retouched flakes. Excavations at Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Shuidonggou shows that although stemmed points are not present at all, a blade technology came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9). Advanced core and flake tools were likely to have been locally developed at 33 ka (SDG Locality 2 and 8), and at 10.8 ka microblade technology appears, but the origin is uncertain (SDG Locality 12) [@yi2014significance; @li2019history]. The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards the Last Glacial Maximum (LGM) [@yi2014significance].

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

Our study of raw material consumption patterns assumes that stone tool was transported and utilized to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

Figure \@ref(fig:volume-density) shows that there is no strong directional temporal pattern of changes in artifact density. The upper bound of artifact density increases in more recent periods. For example, the density of some assemblages are two or three times higher than others, especially after 30 ka. The inset figure shows that assemblages with stemmed points have higher artifact densities than the assemblages without. With the newer tools, hunter-gatherers may have stayed in the same location for longer duration. We may be seeing site occupation patterns in this sample that represent the long term trend of the use of both base camps or residences, represented by the higher artifact densities, and more briefly occupied camps with lower densities. The most striking result here is that maximum size of base camps increases in more recent periods.

@@

Previous work on toolkit composition during the Late Pleistocene of Korea attempted to divide it up into multiple chronological sub-periods. Seong divides five successive assemblage types for the Late Paleolithic in Korea: (1) quartzite and quartz vein artifacts; (2) mostly small quartzite and quartz vein artifacts with some large artifacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30 ka [@seong2015diversity]. Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; 'core tool industry with large flake knapping' during early MIS 3 (59−40 ka), and 'blade industry with tanged point chiefly' during late MIS (40−24 ka) [@lee2012characteristics]. Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement. Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, support the continuation of core tools abundantly throughout the Late Pleistocene. In addition, our data show that side scrapers remained as major tools while other tools occupied only a small portion of the assemblages. As stemmed points and blades increased in the toolkits, both choppers and side scrapers went through minor decreases but still remained abundant.
